---
title: "covid_recovery_analysis"
author: "Cary Ni"
date: "2023-03-23"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(AppliedPredictiveModeling)
library(caret)
library(glmnet)
library(corrplot)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
```

```{r}
# load the external dataset 
load("recovery.rdata")
# change the variable type based on the reference
index_factor = c(3, 4, 5,  9, 10, 13, 14, 15)
index_numer = c(2, 6, 7, 8, 11, 12)
dat[, index_factor]= lapply(dat[, index_factor], as.factor)
# extract 2000 samples for analysis
set.seed(2604)
dat <- dat[sample(1:10000, 2000),]
summary(dat)
# seperate training set and test set
set.seed(2023)
train_row = createDataPartition(y = dat$recovery_time, p = 0.8, list = FALSE)
```

## Exploratory analysis and data visualization

### Visualize potential relationship between reponse variable and numeric predictors

```{r}
# simple visualization of the numeric data
theme1 <- trellis.par.get()
theme1$plot.symbol$col <- rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch <- 16
theme1$plot.line$col <- rgb(.8, .1, .1, 1)
theme1$plot.line$lwd <- 2
theme1$strip.background$col <- rgb(.0, .2, .6, .2)
trellis.par.set(theme1)
featurePlot(x = dat[, index_numer],
            y = dat$recovery_time, 
            plot = "scatter", 
            type = c("p", "smooth"),
            layout = c(3, 2))

```

### Visualize potential relationship between reponse variable and categorical predictors

```{r}
# simple visualization of the categorical data
par(mfrow=c(1,2))
myColors = c(rgb(0.1,0.1,0.7,0.5) , rgb(0.8,0.1,0.3,0.6), rgb(0.8,0.8,0.3,0.5),
             rgb(0.4,0.2,0.3,0.6))

boxplot(recovery_time ~ gender, data = dat, xlab = "gender", col = myColors, ylim=c(0, 150))
boxplot(recovery_time ~ race, data = dat, xlab = "race", col = myColors, ylim=c(0, 150))
boxplot(recovery_time ~ smoking, data = dat, xlab = "smoking status", col = myColors, ylim=c(0, 150))
boxplot(recovery_time ~ hypertension, data = dat, xlab = "hypertension", col = myColors, ylim=c(0, 150))
boxplot(recovery_time ~ diabetes, data = dat, xlab = "diabetes", col = myColors, ylim=c(0, 150))
boxplot(recovery_time ~ vaccine, data = dat, xlab = "vaccine", col = myColors, ylim=c(0, 150))
boxplot(recovery_time ~ severity, data = dat, xlab = "diabetes", col = myColors, ylim=c(0, 150))
boxplot(recovery_time ~ study, data = dat, xlab = "study", col = myColors, ylim=c(0, 150))
```

### Correlation plot to check collinearity between covariates (based on training data) 

```{r}
predictors = model.matrix(recovery_time ~ ., data = dat[train_row, -1])[, -1]
cor(predictors) %>% corrplot(
  method = "circle", type = "full", 
  addCoef.col = 1, number.font =0.5,
  tl.col="black", tl.srt=90, tl.cex = 0.5,
  insig = "blank", diag=FALSE, number.cex = .3)
```

It can be seen from the scatterplots above that `SBP`, `LDL`, and `age` are linearly correlated to recovery time. The correlation plot suggests that `SBP` is positively correlated to `hypertension` while the correlation among `bmi`, `weight`, and `height` are foreseen. In short, there is no obvious collinearity among most of the covariates. 

## Model Training

